FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2025-01-29 manuscript.tex
update 2025-01-29 13:15:59.904362

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8163}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, \& Felix D. Schönbrodt5},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, \& Felix D. Schönbrodt\textsuperscript{5}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# 070e7292.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing.

@@

The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely produce hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate findings that are mutually inconsistent,
as the theory's central claims evade falsification.

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a philosophical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figecs}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
Closer examination of reported deductive research reveals, however, that the link between theory and hypothesis is often tenuous (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit derivation chain from theory to hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely derived from theory,
and test results do not routinely contribute to the improvement of existing theories.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Paul E. Meehl, 1978).

@@

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.
It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind of information is useful.
As a particular FAIR theory evolves, the nature of information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
From this theory, more specific \emph{statistical/mathematical models} could be derived by assuming functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it involves expressing ideas as precise statements.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the precise implementations of decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

To some extent, FAIRness and formalization are orthogonal.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the precision with which they are communicated.
The FAIR principles pertain to theories' documentation, archival and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).

@@

FAIR theory is an example of modular publishing (\textbf{kirczModularityNextForm1998?}).
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
with adequate metadata that is indexed in standardized repositories (\textbf{vandesompelRethinkingScholarlyCommunication2004?}).
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the \href{}{DataCite Metadata Schema} offers a standard way to document the nature of relationships between information artifacts.
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
The present paper considers advantages of Git for FAIR theory.
Git enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).

@@

We will use the empirical cycle as a running example for this tutorial.
The resulting FAIR version of this theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
The empirical cycle is described on page 28 of De Groot and Spiekerman (1969),
and consists of a series of natural language statements:

@@

\begin{figure}
\includegraphics[width=0.33\linewidth]{ec_groot} \includegraphics[width=0.33\linewidth]{ec_wagenmakers} \includegraphics[width=0.33\linewidth]{ec_lissa} \caption{The empirical cycle according to De Groot (left), Wagenmakers and colleagues (center), and Van Lissa (right)}(\#fig:figecs, figures-side)
\end{figure}

@@

\subsection{2. Creating the Project Folder}\label{creating-the-project-folder}

We can write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}, and save it to a new folder.
To help meet the Interoperability and Reusability criteria,
add two more files.
Include a \texttt{README.md} file with instructions for future users of your theory (see: \href{}{What to Include in a README?})
and a \texttt{LICENSE} file with the legal conditions for reuse.
We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work (Ram, 2013; Van Lissa et al., 2021).
To version control our project, we initiate a Git repository in the project folder.
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
Note that the repository
must be set to ``Public'' to take advantage of GitHub's Zenodo integration.
Push the local files to the Git remote repository, and keep them synchronized going forward.

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{fairtheory, meta theory, theory formation, cumulative science, formal models\newline\indent Word count: 8688}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories},
  pdfauthor={Caspar J. Van Lissa1, Aaron Peikert2,3, Andreas M. Brandmaier2,3,4, Felix D. Schönbrodt5, \& Noah N.N. van Dongen6},
  pdflang={en-EN},
  pdfkeywords={fairtheory, meta theory, theory formation, cumulative science, formal models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{FAIR theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories}
\author{Caspar J. Van Lissa\textsuperscript{1}, Aaron Peikert\textsuperscript{2,3}, Andreas M. Brandmaier\textsuperscript{2,3,4}, Felix D. Schönbrodt\textsuperscript{5}, \& Noah N.N. van Dongen\textsuperscript{6}}
\date{}

@@

This is a preprint paper, generated from Git Commit \# c910eaa6.

The authors made the following contributions. Caspar J. Van Lissa: Conceptualization, Formal Analysis, Funding acquisition, Methodology, Project administration, Software, Supervision, Writing -- original draft, Writing -- review \& editing; Aaron Peikert: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Andreas M. Brandmaier: Formal Analysis, Writing -- original draft, Writing -- review \& editing; Felix D. Schönbrodt: Conceptualization, Writing -- review \& editing; Noah N.N. van Dongen: Writing -- review \& editing.

@@

The so-called ``replication crisis'' has prompted extensive reforms in social science (Lavelle, 2021; Scheel, 2022).
Concern that undisclosed flexibility in analyses was a major factor for the abundance of non-replicable findings led to widespread adoption of open science practices like preregistration and replication (Nosek et al., 2015).
These various practices ensure transparent and repeated testing of hypotheses.
However, recent reviews show that most preregistered hypothesis tests are not supported by empirical evidence (Scheel, Schijen, \& Lakens, 2021).
Thus, increased rigor in testing has revealed that the root cause of the replication crisis is more fundamental:
Psychological theories rarely provide hypotheses that are corroborated by evidence.
Furthermore, theories are often so vague that they can accommodate mutually inconsistent findings,
as the theory's central claims evade falsification.

@@

Given these concerns, it is an imbalance that scientific reform initiated by the open science movement has focused primarily on improving deductive methods.
The equally critical inductive processes of theory construction and theory improvement have been largely overlooked\footnote{We use induction to describe inferences from specific observations to general theories; others have coined \emph{abduction} to describe ``inference to the best explanation'' {[}REF Peirce{]}. However, Peirce also acknowledged that abduction ``is not essentially different from induction'' (p.4481). Defining what makes a theory ``best'' requires making auxiliary assumptions. For present purposes, induction and abduction are interchangeable terms.}.
The present paper restores balance by applying, for the first time,
open science principles to psychological theory.
We apply the FAIR principles to scientific theories,
introducing the concept of \emph{FAIR theory} to
facilitate transparent scholarly communication and accelerate cumulative knowledge acquisition.

@@

According to the \emph{empirical cycle} (de Groot, 1961),
a meta-theoretical model of cumulative knowledge acquisition,
research ideally follows a cyclical process with two phases (Figure \ref{fig:figec}).
In the deductive phase, hypotheses derived from theory are tested on data. In the inductive phase, patterns observed in data are generalized to theoretical principles.
In this model, theories are the vehicle of scientists' understanding of phenomena.
Ideally, they are iteratively updated based on deductive testing and inductive theory construction.

\begin{figure}
\centering
\includegraphics{empirical_cycle.pdf}
\caption{\label{fig:figec}A take on the empirical cycle by De Groot}
\end{figure}

In a progressive research program (Lakatos, 1971),
this cycle is regularly completed to iteratively advance our understanding of the studied phenomena
.
There are, however, indications that contemporary psychology falls short of this ideal.
Firstly, because hypothesis-testing research is over-represented in the literature:
According to Kühberger, Fritz, and Scherndl (2014), 89.6\% of papers published in psychology report confirmatory hypothesis tests.
In addition, the link between theory and hypothesis is often tenuous or absent (Oberauer \& Lewandowsky, 2019; Scheel, Tiokhin, Isager, \& Lakens, 2021).
Only 15\% of deductive studies referenced any theory, and theory was often not cited in relation to the hypothesis (McPhetres et al., 2021).
The remaining 85\% of deductive studies lacked an explicit connection between theory and hypothesis.
In the best case, such ungrounded hypotheses are rooted in researchers' implicit theories, in which case it is particularly important to make these explicit Norouzi, Kleinberg, Vermunt, \& Van Lissa (2024).
Or, perhaps the hypotheses are not of substantive interest, such as null hypotheses that exist purely for the purpose of being rejected (Van Lissa et al., 2020), and researchers are simply testing them as part of a cultural ritual (Gigerenzer, Krauss, \& Vitouch, 2004).
Testing ad-hoc hypotheses not grounded in theory does not advance our principled understanding of psychological phenomena.
Put differently: collecting significance statements about ad-hoc hypotheses is much like trying to write novels by collecting sentences from randomly generated letter strings (van Rooij \& Baggio, 2021).

Theory thus has an uncomfortable and paradoxical role in contemporary psychology:
The majority of papers ostensibly test hypotheses,
but these are rarely connected, let alone derived, from theory,
and test results do not routinely reference back to theories, contributing to their improvement or rejection.
The paradoxical role of theory in psychology is perhaps best described by Meehl's observation that theories in psychology ``lack the cumulative character of scientific knowledge. They tend neither to be refuted nor corroborated, but instead merely fade away as people lose interest'' (Paul E. Meehl, 1978).

@@

Since the primary purpose of FAIR theories is to advance scholarly communication about theories,
the method is not contained to any one particular definition.
It is best left to the scholarly community to decide which parts of theory, models, or other aspects should be represented as FAIR theory.
As the practice of FAIRification becomes more embedded,
we expect that it will become increasingly clear what kind and form of information is useful.
As a particular FAIR theory evolves, the nature of the information tracked will likely change.
For example, following Meehl, we could envision a theory that starts out specifying how specific constructs are causally connected.
From this theory, more precise \emph{statistical/mathematical models} could be derived by the theory's suggestions for functional form (e.g., linear effects) and error families (e.g., normal distributions).
This statistical model makes just enough assumptions to allow the estimation of the remaining unknown parameters (e.g., regression slopes) from data.
Then, an even more specific \emph{generative/computational model} could be added, which is completely parameterized (i.e., specific values of regression slopes are also assumed) such that an interpreter (e.g., the R programming language) can use the model to generate new data.
Also, aspects of scientific practice might be added over time, such as commonly used experimental designs (e.g.~longitudinal designs observing change over time), measurement tools (e.g.~different questionnaires used to assess the same construct), or study subjects (e.g.~specific strains of rats).

@@

Concerns about the state of theory in the psychological literature revolve around two issues: theory formalization and theory (re-)use.
Greater formalization increases theories' \emph{empirical content} {[}REF{]} because it expresses ideas as precise statements, clearly demarcating what should (not) be observed.
For example, Baddeley's verbal description of the phonological loop in his theory of working memory allows for at least 144 different implementations depending on the specification of the decay rate, recall success, or rehearsal sequence (Lewandowsky \& Farrell, 2010).
Without committing to specific implementations a-priori,
the theory becomes hard to test.
Committing to specific implementations of the different components, their causal connections, and the functional forms of these relationships makes the theory more precise.
More precise theories are easier to falsify,
which necessitates specific revisions and advances our principled understanding of the phenomena they describe.

Although we are in favor of the advancement towards formal theories,
this might not always be desirable or feasible.
Fortunately, formalization is not required to make theories FAIR.
To some extent, FAIRness and formalization are orthogonal.
FAIR theory imposes no restrictions on the manner in which theories are derived and implemented;
rather, it increases the fidelity and ease with which they are communicated.
The FAIR principles pertain to theories' documentation, archival, and sharing in digital environments, with the aim of enhancing their reusability and extensibility.
For example, a collection of verbal propositions derived through qualitative research could be represented as a FAIR theory.
Conversely, a formal theory is not FAIR if it is represented as a bitmap image in a journal article without any key words to identify it as a theory paper to search engines.
FAIR theory is thus consistent with, but does not require, formalization (also see \emph{Accessibility}).

@@

We propose FAIR theory as an instantiation of modular publishing (\textbf{kirczModularityNextForm1998?}).
At present, the primary unit of social scientific communication is the academic paper.
A paper may depend on multiple resources - materials, data, code, and theory - but these are often merely described in the text.
Modular publishing is the practice of making each of these resources available as independent citable \emph{information artifacts} in their own right,
with adequate metadata that is indexed in standardized repositories (\textbf{vandesompelRethinkingScholarlyCommunication2004?}).
Data sharing is a good example of a modular publishing practice that is widely adopted and increasingly required by funding agencies, journals, and universities.
Scholars can archive information artifacts in repositories like \href{https://zenodo.org/}{Zenodo},
which was developed by \href{https://home.cern/}{CERN} under the European Union's \href{https://www.openaire.eu/}{OpenAIRE} program.
To maintain a persistent record of scholarly communication,
Zenodo mints DOIs for information artifacts - as does, for example, the \href{https://www.crossref.org/}{Crossref} association,
which is used by many academic publishers.
Finally, the \href{}{DataCite Metadata Schema} offers a standard way to document the nature of relationships between information artifacts.
For example, a dataset collected for a specific paper would be archived in Zenodo with the metadata property \texttt{resourceType:\ dataset},
and cross-reference the published paper with \texttt{relationType:\ IsSupplementTo}.
Similarly, FAIR theories can be connected to a specific paper which might serve as the theory's documentation and canonical reference.

@@

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artifacts.
Version control systems, like Git, have long been used to iteratively improve computer code, while managing parallel contributions and allowing for diverging development.
Git tracks line-by-line changes to text-based files,
and maintains a complete history of those changes.
It has long been argued that Git is particularly well-suited to academic work (Ram, 2013).
Git can be used, for example, to facilitate reproducible research, manage distributed collaboration, and improve preregistration (Peikert, Van Lissa, \& Brandmaier, 2021; Van Lissa et al., 2021).
The present paper considers the advantages of Git for FAIR theory.
Git enables explicitly comparing versions of a file (or: theory),
incorporating changes by different authors,
and branching off into different directions (e.g., competing hypotheses) while retaining an explicit link to the common ancestor.
This makes it possible for meta-scientists to study the provenance of a theory and determine how well different versions of a theory explain empirical evidence (Van Lissa, 2023).

@@

Given that we structured our argument around the importance of FAIR theory for cumulative knowledge production through scientific research around the \emph{empirical cycle},
we decided to use it as an example for this tutorial.
The resulting FAIR theory is available at \url{https://doi.org/10.5281/zenodo.14552329}.
The empirical cycle is described on page 28 of De Groot and Spiekerman (1969).
Note that, while De Groot does not explicitly refer to the empirical cycle as a ``theory'', he derives it from ``a theory of thinking''.
We can thus consider it a meta-theory of theory construction.
The ``empirical cycle'' theory presented by De Groot consists of a series of natural language statements:

@@

@@

We can now write this implementation of the empirical cycle to a text file, say \texttt{empirical\_cycle.dot}.

\subsection{2. Creating a Project Folder}\label{creating-a-project-folder}

Create a new folder and copy the theory file from the previous step into it.
To help meet the Interoperability and Reusability criteria,
add two more files:
A README.md file with instructions for future users of your theory,
and a LICENSE file with the legal conditions for reuse.
We recommend the \texttt{CC0} license, but other options are available, see \href{https://choosealicense.com/non-software/}{https://choosealicense.com}.

\subsubsection{What's in a README?}\label{whats-in-a-readme}

The readme should contain information to help people get started with using your FAIR theory.
We suggest the following elements:

\begin{itemize}
\tightlist
\item
  Title, prefaced with \texttt{\#\ FAIR\ theory:\ The\ Theory\textquotesingle{}s\ Name}
\item
  Description: A plain-text description of the theory and its scope
\item
  Interoperability: Most README files contain a section labeled ``Getting Started'', ``Instructions'', or ``How to Use''. From a FAIR perspective, such a section might be better labeled ``Interoperability'', or ``How to Use (Interoperability)''. We propose explicitly addressing the theory's X-interoperability, telling users exactly what they can use the theory for, and how. For example, our example is implemented in the DOT language for describing graphs, so we would could provide instructions here on how to plot a DOT graph.
\item
  Contributing: Pertaining to the Reusability criterion, this section should tell users the \emph{social expectations regarding reuse and contributions}.
\item
  License: The legal complement to the preceding section, this section should refer readers to the LICENSE file to learn about the \emph{legal conditions of reuse}.
\item
  Citing this work: Tell users how to cite the theory. Note that this section is redundant with the Zenodo archive, which has a preferred citation field. The disadvantage of redundant information is that you may have to maintain this section of the README going forward. The advantage is that documenting related works in the README makes it more readily accessible to users. We suggest a compromise: to retain this section, but refer the reader to the Zenodo page.
\item
  Related works: This section should refer to the work that the FAIR theory is derived from, or documented in. Again, this is redundant with metadata entered in Zenodo (step 5). We nevertheless recommend using this section to refer to Zenodo, and/or to document one canonical reference for the theory that is unlikely to change going forward. For example, we referenced the original empirical cycle paper here:
\end{itemize}

\begin{verbatim}
This repository contains an implementation of the "empirical cycle",
a model proposed by De Groot and Spiekerman (1969, p. 28). See Zenodo for other related works.

> De Groot, A. D., & Spiekerman, J. A. A. (1969). Methodology:
Foundations of inference and research in the behavioral sciences.
De Gruyter Mouton. https://doi.org/10.1515/9783112313121
\end{verbatim}

@@

The field of computer science provides well-established processes for creating information artefacts that can be iteratively improved.
In particular, the practice of version control offers extensive benefits for scientific work (Ram, 2013; Van Lissa et al., 2021).
To version control our project, we initiate a Git repository in the project folder.
We subsequently create a remote repository to host a copy of this local Git repository on GitHub, which will in turn be archived.
Note that the repository
must be set to ``Public'' to take advantage of GitHub's Zenodo integration.

Push the local files to the Git remote repository, and keep them synchronized going forward.
